---
title: "Palmer Penguin Analyses"
output:
  html_document:
    toc: true
    toc_depth: 2          
    toc_float: true       
    code_folding: hide
date: "2024-12-09"
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# Initialize renv Environment to install required packages 
renv::restore()
```

### Load the packages

```{r, message = FALSE, warning = FALSE}
library(palmerpenguins)
library(ggplot2)
library(janitor)
library(dplyr)
library(here)
library(svglite)
library(stats)
library(ggsignif)
library(broom)
library(knitr)
library(kableExtra)
library(renv)
library(ggpubr)
library(car)
library(rstatix)
library(caret)
```

```{r, include=FALSE}
#Set location as here
here()
```

### Call the required functions

```{r, include=TRUE}
source(here("Functions/Cleaning.R"))
source(here("Functions/AssumptionCheck.R"))
source(here("Functions/SavingFigure.R"))
source(here("Functions/Summary.R"))
source(here("Functions/Table.R"))
```

# **Section 1:** Data Visualization for Science Communication

```{r echo=FALSE, warning=FALSE, fig.align='center', message = FALSE}

#Filter the data to remove extreme values 
penguins_filtered <- penguins %>%
  filter(
    `bill_length_mm` > 33,
    `bill_depth_mm` < 30
  )

ggplot(penguins_filtered, aes(x = `bill_depth_mm`, y = `bill_length_mm`, shape = sex, color = sex)) +
  geom_point(size = 5, alpha = 1.0) +  # Add points
  geom_smooth(method = "lm", se = FALSE, color = "blue", aes(group = 1)) +  # Single trend line 
  labs(x = "Culmen depth", y = "length", title = "Comparing lengths") +  # Add axis labels and title
  theme(
    axis.title = element_text(size = 5),  # Adjust size of axis titles
    plot.title = element_text(size = 12),  # Adjust size of the plot title
    legend.text = element_text(size = 5),  # Adjust size of legend text
    legend.title = element_text(size = 10) # Adjust size of legend title
  )+
  scale_y_continuous(limits = c(30, 55)) +  # Shrink the y-axis range
  scale_x_continuous(limits = c(10, 25))  # Shrink the x-axis range

```

### b) Write about how your design choices mislead the reader about the underlying data (200-300 words).

The presented figure employs multiple design features that mislead the viewer and misrepresent the underlying data, reflecting common issues in scientific reporting and reproducibility. A nature survey found that 70% of researchers have tried and failed to reproduce another scientist's work, and more than half have failed to reproduce their own [@baker2016reproducibility]. The lack of reproducibility is driven by factors such as the pressure to publish, which prioritises novel findings over rigour, and selective reporting, which biases the literature toward positive results [@baker2016reproducibility].

The figure above is a good example of misleading data representation, showing a negative correlation between culmen length and depth. However, closer analysis (e.g., colouring dots by species) would reveal a positive correlation within each species. This is known as Simpson's paradox. This data was also filtered to exclude extreme values, a practice known as cherry-picking, which can obscure the true relationship. Both cherry-picking and Simpson's paradox are common issues in closed research processes. This figure has multiple aesthetic problems which interfere with interpretation. The text is small and ambiguous, lacking units and specificity in the measurements. The data points are large and overlapping, making it hard to understand their true distribution. Finally, the scale does not start at zero, exaggerating the negative slope, and the data points are clumped in the middle of the graph. Overall, this makes it hard to interpret the graph, and what they can gather is misleading.

The push for open access, open data, open source, and other open scholarship practices should prevent the publication of ambiguous, misleading graphs like the one above. The potential benefits of open science, such as increased media attention and collaboration opportunities, are driving its growing popularity. However, insufficient public evidence of open publishing's effectiveness currently limits full compliance among scientists [@mckiernan2016open].

------------------------------------------------------------------------

# **Section 2:** Data Pipeline

# [Introduction]{.underline}

> Gentoo, Adelie, and Chinstrap are three species of penguin that are found in the Palmer Archipelago near Antarctica. This analysis seeks to determine whether these species occupy different feeding niches by comparing their bill morphology, specifically the culmen depth. Morphological plasticity in bill size has been shown to relate closely to diet and thus, niche differentiation [@laranjeiro2022variation]. Specifically, culmen depth plays a crucial role in determining the strength and functionality of a penguin's bill, directly influencing the types of prey the species can access. A deeper bill typically allows for the capture and handling of larger, more robust prey. As a result, if penguins have similar culmen depths, it can be inferred that they share similar feeding niches. This study aims to compare the culmen depths among the three penguin species to gain insights into their specific feeding niches.

:::: hypothesis
::: {style="border: 2px solid #000; padding: 10px; background-color: #f9f9f9; border-radius: 5px;"}
**H0:** There is no significant difference in the mean culmen depth among the penguin species.

**H1:** At least one Penguin species mean is significantly different from the others.
:::
::::

### Viewing the raw data

```{r, include=TRUE}
kable(head(penguins_raw), caption = "Preview of Penguin Dataset")
```

### Cleaning and filtering the data

```{r, include=TRUE}

#Save a raw copy of the data before cleaning it. 
write.csv(penguins_raw, "data/penguins_raw.csv")

#Clean the data 
penguins_clean <- Cleaning(penguins_raw) #See annotation within the cleaning function

#Save the cleaned data set 
write.csv(penguins_clean, "data/penguins_clean.csv")

#Select the collumn that will be analysed (culmen_depth_mm) and remove rows with NA. This ensures NA are only removed from the collumn being used. 
penguins_clean_subset <- penguins_clean %>% select(culmen_depth_mm, species) %>% remove_NA() #Remove_NA is from within the cleaning package. 

```

### Visualizing the data

```{r echo=TRUE, warning=FALSE, fig.align='center', message = FALSE}

# Plot a histogram of all the raw data 
Flipper.Histogram <- ggplot(penguins_clean_subset, aes(x = culmen_depth_mm, fill = species)) +  # Load ggplot2 and prepare the plot
  geom_histogram(alpha = 0.6, position = "identity", binwidth = 0.3) +  # Add a histogram layer with transparency and overlapping histogram bars 
  labs(
    title = "Culmen Depth Distribution of the Palma Pengiuns",
    x = "Culmen Depth Length (mm)",
    y = "Frequency",
    caption = "Figure 1: Histogram showing the distribution of culmen depth measurements (in mm) for different penguin \n species, highlighting the similarities and variations in culmen depth between species") + # Add labels for the title and axes
  theme(
    plot.caption = element_text(hjust = 0, size = 9) # Align the caption to the left and size it
  ) + theme_bw() #Set the theme 

# Display the plot
print(Flipper.Histogram)

```

```{r, message=FALSE, warning=FALSE}
#Saving the figure: Apply the save function
save_flipper_plot_svg(Flipper.Histogram,
                      here("Figures", "Flipper.Histogram.svg"), size = 25, scaling = 1.25)
```

### Summary statistics

```{r}
#Run the summary function to get the mean and SD for culmen depth
summarize_data(penguins_clean_subset, "species", "culmen_depth_mm", "Table 1: Summary Statistics of Beak Depth (mm) by Species") #Function details in function file 
                
```

### Insight from exploration and cleaning

> The data, sourced from the palmerpenguin R package, includes various measurements collected from 344 penguins, representing three species: Adelie, Gentoo, and Chinstrap. The exploratory plot and summary statistics reveal that Adelie and Chinstrap penguins have similar culmen lengths, while Gentoo penguins exhibit greater variation. Statistical analysis is required to assess the significance of these differences.

# [Methods - statistical analyses]{.underline}

> A fixed linear model was created, with species as the explanatory variable and culmen depth as the response variable. Residual and QQ plots were utilised to evaluate model assumptions, revealing that the data satisfied the homoscedasticity assumption (confirmed by Levene's test) but violated the normality assumption (confirmed by the Shapiro-Wilk test). However, the QQ plot showed only minor deviations from normality, and given the large sample size (344), linear models are generally robust to such slight violations, so the data was not transformed. A one-way ANOVA was then performed to assess whether the mean culmen lengths between species were significantly different, with significance determined at α=0.05. Finally, a pairwise comparison with a Bonferroni adjustment, accounting for Type I error, was performed to confirm the pairwise differences.

### 1) Linear model 1

```{r, include=TRUE}
# Generate the linear model 
Culmen_model <- lm(culmen_depth_mm ~ species, data = penguins_clean_subset) 
```

### Checking assumptions: graphical assessment

The residual plot shows there is no violation of homoscedasticity in the residuals assumptions while the QQ plot shows minor normality violations. I will run a statistical test to check whether this violation is significant.

```{r echo=TRUE, warning=FALSE, fig.align='center', message = FALSE}

#Apply the Assumption checking function
Assumption_check(Culmen_model, "Figure 2: The evenness of the residual spread in the three clusters (species) suggests no \n violation of homoscedasticity.\n Figure 3: Residuals mostly follow the diagonal, with slight tail deviations suggesting minor normality \n  violations.") #Function requires the model and the caption to be specified. #Function details in function file 
```

### Checking assumptions: statistical assessment

The homoscedasticity of the data was confirmed by The Levene's test (P \> 0.05). However, the Shapiro-Wilk test identified a statistically significant violation of normality (P \< 0.05). Therefore, I will apply a data transformation.

```{r, include=TRUE}

#Perform statistical tests and create a table 
results_table <- list(
  Shapiro_Wilk = shapiro.test(residuals(Culmen_model)), # Shapiro-Wilk Test for normality 
  Levene = leveneTest(culmen_depth_mm ~ factor(species), data = penguins_clean_subset) # Levene's test for homogeneity of variance
) %>%
  {
    data.frame( # Create a data frame with the results
      Test = c("Shapiro-Wilk", "Levene's Test"),
      Statistic = c(.$Shapiro_Wilk$statistic, .$Levene$`F value`[1]), # Extract test statistics
      P_Value = c(.$Shapiro_Wilk$p.value, .$Levene$`Pr(>F)`[1]) # Extract P-values
    )
  } %>%
  mutate(
    Statistic = round(Statistic, 3), # Round test statistics to 3 decimal places
    P_Value = round(P_Value, 3)      # Round P-values to 3 decimal places
  )

#Apply function to display the results table 
create_styled_table(results_table, c("Test", "Test Statistic", "P-Value"), "Table 3: Test Results for Shapiro-Wilk and Levene's Test") #Function details in function file 

```

### 2) Linear model 2

I performed a Box Cox transformation and created a second linear model.

```{r, include=TRUE}
# Set up the Box Cox transformation
bc_trans <- BoxCoxTrans(penguins_clean_subset$culmen_depth_mm) 

# Apply the Box-Cox transformation to the Depth data and create a new column with the transformed data
Penguins_new <- penguins_clean_subset %>%
  mutate(TransformedDepth = predict(bc_trans, penguins_clean_subset$culmen_depth_mm)) 

#Re-fit the linear model 
Culmen_model_2 <- lm(TransformedDepth ~ species, data = Penguins_new) 

```

### Re-Checking assumptions: statistical assessment

The test results confirm that the normality and homoscedasticity assumptions are met as the results are insignificant (P\>0.05).

```{r, include=TRUE}

#Perform statistical tests and create a table 
results_table <- list(
  Shapiro_Wilk = shapiro.test(residuals(Culmen_model_2)), # Shapiro-Wilk Test for normality 
  Levene = leveneTest(culmen_depth_mm ~ factor(species), data = Penguins_new) # Levene's test for homogeneity of variance
) %>%
  {
    data.frame( # Create a data frame with the results
      Test = c("Shapiro-Wilk", "Levene's Test"),
      Statistic = c(.$Shapiro_Wilk$statistic, .$Levene$`F value`[1]), # Extract test statistics
      P_Value = c(.$Shapiro_Wilk$p.value, .$Levene$`Pr(>F)`[1]) # Extract P-values
    )
  } %>%
  mutate(
    Statistic = round(Statistic, 3), # Round test statistics to 3 decimal places
    P_Value = round(P_Value, 3)      # Round P-values to 3 decimal places
  )

#Apply function to display the results table 
create_styled_table(results_table, c("Test", "Test Statistic", "P-Value"), "Table 3: Test Results for Shapiro-Wilk and Levene's Test") #Functions details in function file 

```

### 3) One- Way ANOVA

```{r, include=TRUE}
# Perform the anova and format the results 
anova_table <- Anova(Culmen_model_2) %>% # Perform the ANOVA
  as.data.frame() %>% # Turn the results into a data frame
  setNames(c("Sum Sq", "Df", "F Value", "P-Value")) %>% # Set the column names
  mutate(
    `P-Value` = formatC(`P-Value`, format = "e", digits = 3) # Format P-values in standard form (or will present as 0)
  )


#Apply function to display the results table 
create_styled_table(anova_table, c("Sum of Squares", "Degrees of Freedom", "F Value", "P-Value"), "Table 3: ANOVA Results") #Function details in function folder 


```

### 4) Pairwise comparison

```{r, include=TRUE}

#Perform a pwc and generate a table of results 
pwc <- penguins_clean_subset %>% # Specify the data set
  pairwise_t_test( # Specify the test
    culmen_depth_mm ~ species, paired = FALSE, # Specify the variables
    p.adjust.method = "bonferroni" # Apply the Bonferroni adjustment
  ) 

pwc_table <- pwc %>%  select(group1, group2, p, p.adj) %>% # Select the information to present
  mutate(
    p = formatC(p, format = "e", digits = 3),       # Format p-values 
    p.adj = formatC(p.adj, format = "e", digits = 3) # Format adjusted p-values 
  )

#Apply function to display the results table 
create_styled_table(pwc_table, c("Group 1", "Group 2", "P-Value", "Adjusted P-Value"), "Table 4: Pairwise Comparisons with Bonferroni Adjustment") #Function details in function folder 


```

# [Results and discussion]{.underline}

> The results of the one-way ANOVA (F(2, 339) = 259.79, p \< 0.001), confirmed at least one species of Penguin had a significantly different culmen depth, so the null hypothesis can be rejected. The pairwise comparison revealed that there’s no significant difference between Adelie and Chinstrap (P\>0.05), while both Adelie and Chinstrap significantly differ from Gentoo (P\<0.0001). These results are displayed in Figure 4.

> From these results it appears Adelie and Chinstrap and more likely to prey on similar species and thus inhabit a similar feeding niche compared to Gentoo. These findings correlate with what we know about penguin feeding ecology. The Adelie and Chinstrap penguin have more similar diets (eat primarily krill), than the Gentoo penguin which is more generalist (eat a mixture of crustaceans, small fish, and squid).

> However, it is important to note that while culmen depth can provide insight into the general foraging behaviour and feeding niche, the culmen length may also be a crucial factor in shaping their feeding ecology. Differences in culmen length could reveal additional patterns in prey selection and feeding behaviour, potentially indicating that all three species inhabit slightly different feeding niches. Other factors like body size will also contribute to prey selection. If considered in isolation, these results may imply that Adelie and Chinstrap occupy exactly the same feeding niche. Further analysis is required to compare other morphological features like culmen length and body size between the three species.

```{r echo=TRUE, warning=FALSE, fig.align='center', message = FALSE}

# Create a box plot including the results of the ANOVA and the pwc
Results.figure <- ggplot(data = penguins_clean_subset, aes(x = penguins_clean_subset$species, y = culmen_depth_mm)) +  # Load ggplot2 and prepare the plot
  geom_boxplot(aes(fill = species), alpha = 0.7, width = 0.5) +  # Add box plot layer
  labs(
    y = "Culmen Depth (mm)", 
    x = "Species", 
    title = "Comparing the Culmen Depth of the Palma Penguins",
    subtitle = "ANOVA, F(2, 339) = 359.79, p < 0.0001", 
    caption = "Figure 4: Box plot comparing the culmen depth of the 3 species of Palma Penguin.  Includes the results of the \n ANOVA which are significant ( P<0.0001) and the Pairwise comparisons with Bonferroni correction"
  ) +  # Labeling 
  stat_pvalue_manual(pwc, y.position = c(23, 24, 25)) +  # Add pwc results 
  scale_y_continuous(limits = c(13, 25)) +  # Position pwc lines 
  theme_bw() +  # Set the theme 
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 9),
    plot.caption = element_text(size = 10), 
    legend.position = "none"  # Edit font sizes and positions
  )

print(Results.figure)
  
```

```{r, message=FALSE, warning=FALSE}
#Saving the figure: Apply the save function 
save_flipper_plot_svg(Results.figure,
                      here("Figures", "Results.figure.svg"), size =25, scaling = 1.25)
```

# [Conclusion]{.underline}

> In conclusion, the null hypothesis of no difference between the three culmen depths can be rejected. While the Adelie and Chinstrap penguins exhibit similar culmen depths, the Gentoo penguins demonstrate a significantly different culmen depth, and these patterns inform us about their feeding niches. However, further analyses of other traits affecting the prey available are required to further understand their feeding niche. Even so, this is an interesting example of how morphological divergence is involved in niche specialization within genera, reducing competition and allowing co-existence. Different combinations of penguin species are found on Torgersen, Dream, and Biscoe Islands; however, all three species never coexist on the same island. It would be interesting to investigate whether the presence of different species on an island influences their unique ecological specializations.

# [References]{.underline}

::: {#refs}
:::

------------------------------------------------------------------------

```{r, include = FALSE, message=FALSE, warning=FALSE}
#Setting up library store 
renv::init()
```

```{r, include = FALSE, message=FALSE, warning=FALSE}
#Run this code after editing the libraries to save changes 
renv::snapshot()
```

# **Section 3:** Open Science

### a) GitHub

*GitHub link:* <https://github.com/Anonymous94394/PalmerPenguins_1.git>

### b) Share your repo with a partner, download, and try to run their data pipeline.

*Partner's GitHub link:*

*You **must** provide this so I can verify there is no plagiarism between you and your partner.*

### c) Reflect on your experience running their code. (300-500 words)

-   *What elements of your partner's code helped you to understand their data pipeline?*

-   *Did it run? Did you need to fix anything?*

-   *What suggestions would you make for improving their code to make it more understandable or reproducible, and why?*

-   *If you needed to alter your partner's figure using their code, do you think that would be easy or difficult, and why?*

### d) Reflect on your own code based on your experience with your partner's code and their review of yours. (300-500 words)

-   *What improvements did they suggest, and do you agree?*

-   *What did you learn about writing code for other people?* ::::
